---
title: "Cyclistic's Case Study"
author: "Juliano Zen"
date: "23/02/2022"
output:
  pdf_document: default
  html_document: default
---

### About the Company:
"A bike-share program that features more than 5,800 bicycles and 600 docking stations. 
Cyclistic sets itself apart by also offering reclining bikes, hand tricycles, and cargo bikes, making bike-share more inclusive to people with disabilities and riders who can’t use a standard two-wheeled bike. 
The majority of riders opt for traditional bikes; about 8% of riders use the assistive options. Cyclistic users are more likely to ride for leisure, but about 30% use them to commute to work each day." - From Coursera's Case Study Project.


### Project overview:

* Cyclistic's marketing goal; to convert casual riders to Cyclistic members (annual membership).

* Key task for this analysis - How do annual members and casual riders use Cyclistic bikes
differently.


##### Libraries setup for R project:

Run packages installations if not already installed.

```{R Packages, include=FALSE}
install.packages("tidyverse", repos = "http://cran.us.r-project.org") #helps wrangle data
install.packages("lubridate", repos = "http://cran.us.r-project.org") #helps wrangle date attributes
```

Then load libraries up.

```{R Libraries}
library(tidyverse)
library(lubridate)
library(dplyr)
```

##### Data collection and transformation:

Data collected from Feb/2021 to Jan/2022. == Data made available by Motivate International Inc. under this [license](https://ride.divvybikes.com/data-license-agreement).

Data sets:

```{R Datasets}
"m02_2021" <- read_csv("202102-divvy-tripdata.csv")
"m03_2021" <- read_csv("202103-divvy-tripdata.csv")
"m04_2021" <- read_csv("202104-divvy-tripdata.csv")
"m05_2021" <- read_csv("202105-divvy-tripdata.csv")
"m06_2021" <- read_csv("202106-divvy-tripdata.csv")
"m07_2021" <- read_csv("202107-divvy-tripdata.csv")
"m08_2021" <- read_csv("202108-divvy-tripdata.csv")
"m09_2021" <- read_csv("202109-divvy-tripdata.csv")
"m10_2021" <- read_csv("202110-divvy-tripdata.csv")
"m11_2021" <- read_csv("202111-divvy-tripdata.csv")
"m12_2021" <- read_csv("202112-divvy-tripdata.csv")
"m01_2022" <- read_csv("202201-divvy-tripdata.csv")
```

Format checking: Compare column names prior to joining data sets.

```{R Collumn consistency comparison}
colnames(m01_2022)
colnames(m02_2021)
colnames(m03_2021)
colnames(m04_2021)
colnames(m05_2021)
colnames(m06_2021)
colnames(m07_2021)
colnames(m08_2021)
colnames(m09_2021)
colnames(m10_2021)
colnames(m11_2021)
colnames(m12_2021)
```

Column names, check.
No renaming needed.

Incongruity checking:

```{R}
str(m01_2022)
str(m02_2021)
str(m03_2021)
str(m04_2021)
str(m05_2021)
str(m06_2021)
str(m07_2021)
str(m08_2021)
str(m09_2021)
str(m10_2021)
str(m11_2021)
str(m12_2021)
```
All columns assigned correctly. Check
No further action required.

Combining checked data frames: !!! **WARNING, chunk crashing R. Run syntax in console.** !!!

```{r Combining data}
all_trips <- bind_rows(m02_2021, m03_2021, m04_2021, m05_2021, m06_2021, m07_2021, m08_2021, m09_2021, m10_2021, m11_2021, m12_2021, m01_2022)
```


Adding columns: Date, month, day, year, day of week.

```{r, Dates}
all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

Adding ride length calculation:

```{r Ride Length}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
```

Converting ride length from factor to numeric:

```{r}
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

Cleaning bad data, getting rid of values for ride length less than zero: !!! **WARNING, chunk crashing R. Run syntax in console.** !!

```{r}
all_trips_v2 <- all_trips[!(all_trips$ride_length<0),]
```

Check for NA's in ride_length:

```{r}
summary(all_trips_v2$ride_length)
```

Removing NA's if needed: Wasn't the case, no NA's found. 
**Code referenced for future uses**
all_trips_v3 <- all_trips_v2[complete.cases(all_trips_v2),] #Creates a new DF without NA's


##### Descriptive analysis 

```{r}
mean(all_trips_v2$ride_length) #straight average (total ride length / rides)
median(all_trips_v2$ride_length) #midpoint number in the ascending array of ride lengths
max(all_trips_v2$ride_length) #longest ride
min(all_trips_v2$ride_length) #shortest ride
```

Summary:

```{r}
summary(all_trips_v2$ride_length)
```

Casual users x members comparison:

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```

Average ride time by each day for members vs casual users:

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

Average ride time by each day for members vs causal users (ordered):

```{r}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("domingo", "segunda-feira", "terça-feira", "quarta-feira", "quinta-feira", "qexta-feira", "sábado"))
```

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```


```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%                  #groups by usertype and weekday
  summarise(number_of_rides = n()                       #calculates the number of rides and average duration 
  ,average_duration = mean(ride_length)) %>%            # calculates the average duration
  arrange(member_casual, weekday)                       # sorts
```

##### Number of rides per Rider Type:

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```

#### Average trip duration:

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")
```

##### File exported for further analysis: 

avg_ride_length.csv | counts data frame

```{r, include=FALSE }
counts <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
write.csv(counts, file = "C:\\Users\\T-Gamer\\Documents\\R\\Case Studies\\Cyclistic\\avg_ride_length.csv")
```

all_trips_v2 | full data frame

```{r, include=FALSE }
write.csv(all_trips_v2, file = "R\\Case Studies\\Cyclistic\\all_trips_v2.csv")
```
